\documentclass[11pt,]{article}
\usepackage[margin=1in]{geometry}
\newcommand*{\authorfont}{\fontfamily{phv}\selectfont}
\usepackage[]{mathpazo}
\usepackage{abstract}
\renewcommand{\abstractname}{}    % clear the title
\renewcommand{\absnamepos}{empty} % originally center
\newcommand{\blankline}{\quad\pagebreak[2]}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}} 
\usepackage{longtable,booktabs}

\usepackage{parskip}
\usepackage{titlesec}
\titlespacing\section{0pt}{12pt plus 4pt minus 2pt}{6pt plus 2pt minus 2pt}
\titlespacing\subsection{0pt}{12pt plus 4pt minus 2pt}{6pt plus 2pt minus 2pt}

\titleformat*{\subsubsection}{\normalsize\itshape}

\usepackage{titling}
\setlength{\droptitle}{-.25cm}

%\setlength{\parindent}{0pt}
%\setlength{\parskip}{6pt plus 2pt minus 1pt}
%\setlength{\emergencystretch}{3em}  % prevent overfull lines 

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{lastpage}
\renewcommand{\headrulewidth}{0.3pt}
\renewcommand{\footrulewidth}{0.0pt} 
\lhead{}
\chead{}
\rhead{\footnotesize Econ/HLTH 470: Research in Health
Economics -- 1/10/2023 - 4/24/2023}
\lfoot{}
\cfoot{\small \thepage/\pageref*{LastPage}}
\rfoot{}

\fancypagestyle{firststyle}
{
\renewcommand{\headrulewidth}{0pt}%
   \fancyhf{}
   \fancyfoot[C]{\small \thepage/\pageref*{LastPage}}
}

%\def\labelitemi{--}
%\usepackage{enumitem}
%\setitemize[0]{leftmargin=25pt}
%\setenumerate[0]{leftmargin=25pt}




\makeatletter
\@ifpackageloaded{hyperref}{}{%
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
}
\@ifpackageloaded{color}{
    \PassOptionsToPackage{usenames,dvipsnames}{color}
}{%
    \usepackage[usenames,dvipsnames]{color}
}
\makeatother
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={ ()},
             pdfkeywords = {},  
            pdftitle={Econ/HLTH 470: Research in Health Economics},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls


\setcounter{secnumdepth}{0}

\usepackage{longtable}




\usepackage{setspace}

\title{Econ/HLTH 470: Research in Health Economics}
\author{Ian McCarthy}
\date{1/10/2023 - 4/24/2023}


\begin{document}  

		\maketitle
		
	
		\thispagestyle{firststyle}

%	\thispagestyle{empty}


	\noindent \begin{tabular*}{\textwidth}{ @{\extracolsep{\fill}} lr @{\extracolsep{\fill}}}


E-mail: \texttt{\href{mailto:ian.mccarthy@emory.edu}{\nolinkurl{ian.mccarthy@emory.edu}}} & Web: \href{http://econ470s23.classes.ianmccarthyecon.com}{\tt econ470s23.classes.ianmccarthyecon.com}\\
Office Hours: MW 1:30pm-3:00pm  &  Class Hours: MW 10:00am-11:15am\\
Office: PAIS 573  & Class Room: Rich 211\\
	&  \\
	\hline
	\end{tabular*}
	
\vspace{2mm}
	


\hypertarget{course-description}{%
\section{Course Description}\label{course-description}}

This is a capstone course that combines health economics and human
health content with econometrics data science. The course is therefore
heavily applied in nature. We will work through four ``canonical''
research designs, in each case using real world data. You will also
complete an empirical research project using real world data and employ
econometric methods to analyze a research question relevant to
contemporary health care issues and/or health policy.

\textbf{Prerequisites:} Econ 320 (Econometrics) and at least one of the
following health economics and policy classes: HLTH 370, Econ 371, Econ
372.

\hypertarget{learning-outcomes}{%
\section{Learning Outcomes}\label{learning-outcomes}}

The content of the course is split into three general areas: 1) data
management in the real world; 2) health policy and health care
institutions; and 3) empirical methods in program evaluation and causal
inference. Each area of the course will be covered by way of an example
research question, which you can use to guide your own projects
throughout the semester. By the end of this course, you will be able to:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Organize project files using Git and GitHub
\item
  Clean and manage several datasets using tidy data in RStudio
\item
  Summarize and visualize data with RStudio and the ggplot2 package
\item
  Implement selected methods for causal inference using real data
\item
  Explain research results with a written report and discussion
\end{enumerate}

There are more specific learning outcomes for each module described on
the relevant module's page of our class website, accessible
\href{https://econ470s23.classes.ianmccarthyecon.com/}{here}.

\hypertarget{text-software-and-class-materials}{%
\section{Text, Software, and Class
Materials}\label{text-software-and-class-materials}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Where to go:} The best place for information on the class is
  our class website,
  \href{https://econ470s23.classes.ianmccarthyecon.com/}{here}. I use
  the website as an easier way for you to access course materials,
  including notes/slides, a detailed schedule, and assignments. I'll
  also use Canvas to distribute any papers or specific readings that
  aren't available publicly, as well as post information that I don't
  want public like virtual meeting passwords (if necessary) and grades.
\item
  \textbf{Readings:} There is no required textbook for this class, but
  there are several books that I recommend for additional information
  and examples. The first is not free but is still very good:
\end{enumerate}

\begin{itemize}
\tightlist
\item
  J. Angrist and J. Pischke \emph{Mostly {Harmless} {Econometrics}}
  (Princeton, NJ: Princeton University Press, 2009).
\end{itemize}

The following are also excellent resources, with the added bonus of
having freely available versions of these texts online. Here are links
to the free online versions:

\begin{itemize}
\tightlist
\item
  \href{https://r4ds.had.co.nz/}{R for Data Science}
\item
  \href{https://mixtape.scunning.com/}{Causal Inference: The Mixtape}
\item
  \href{https://theeffectbook.net/}{The Effect: An Introduction to
  Research Design and Causality}
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\item
  \textbf{Statistics Software:} I'll use \texttt{R} for my work and in
  all my answer keys and examples. I encourage you to use \texttt{R} as
  well, since that will be the language for which I can help you the
  most with any questions. That said, if you can answer all of the
  questions correctly with another software, feel free to do so. I am
  well-versed in \texttt{Stata} but less so in \texttt{Python}. If you
  choose to use \texttt{Python} (not a bad choice given its popularity),
  then I can certainly help with general questions and concepts, but I
  will be less helpful with any syntax issues.
\item
  \textbf{Accessing Data:} For all assignments and in-class activities,
  the necessary raw data will be available in our class OneDrive folder.
  For transparency, I will also point you to other GitHub repositories
  where available. We'll spend some time on Git and GitHub at the
  beginning of the class, so don't get too worried if this is new to
  you.
\item
  \textbf{Logistics:} For day-to-day communication and announcements,
  I'll use \emph{Canvas}. Please be sure you've set up Canvas alerts so
  that you see all of these announcements in a timely manner.
\item
  \textbf{Turning things in:} All assignments should be submitted as
  links to GitHub repositories in \emph{Canvas}.
\end{enumerate}

\hypertarget{course-policies}{%
\section{Course Policies}\label{course-policies}}

Various policies for this course are described below. Basically, let's
all work to be good citizens and take seriously our various roles as a
student, teacher, friend, colleague, human, etc.

\hypertarget{class-meetings}{%
\subsection{Class meetings}\label{class-meetings}}

All regular class meetings will take place in the \textbf{Rich Memorial
Building, Room 211} on Mondays and Wednesdays from 10:00am to 11:15am.

\hypertarget{office-hours}{%
\subsection{Office hours}\label{office-hours}}

My designated office hours will be on Monday and Wednesday from 1:30pm
to 3:00pm in PAIS 573. While we call these ``office hours'' they really
should be ``student hours''. These are set times dedicated to all of
you. This means that I will be in my office waiting for you to come by
with whatever questions you have. I'm also available outside of those
designated times. I just ask that you schedule an appointment in
advance. \href{https://mccarthy-meetings.youcanbook.me/}{Make an
appointment with me here}. Unless we arrange otherwise, my office hours
will be exclusively in-person.

For any questions that you don't think need a full meeting, you can also
always reach out to me through email. I do my best to respond within 24
hours to all emails.

\hypertarget{academic-integrity}{%
\subsection{Academic integrity}\label{academic-integrity}}

The Honor Code is in effect throughout the semester. By taking this
course, you affirm that it is a violation of the code to cheat on exams,
to plagiarize, to deviate from the teacher's instructions about
collaboration on work that is submitted for grades, to give false
information to a faculty member, and to undertake any other form of
academic misconduct. You agree that the instructor is entitled to move
you to another seat during examinations, without explanation. You also
affirm that if you witness others violating the code you have a duty to
report them to the honor council. Students who violate the Honor Code
may be subject to a written mark on their record, failure of the course,
suspension, permanent dismissal, or a combination of these and other
sanctions. The Honor Code may be reviewed at:
\href{http://catalog.college.emory.edu/academic/policies-regulations/honor-code.html}{Emory
Honor Code Policies}.

\hypertarget{accessibility-services}{%
\subsection{Accessibility services}\label{accessibility-services}}

If you anticipate issues related to the format or requirements of this
course, please meet with me. I would like us to discuss ways to ensure
your full participation in the course. If you determine that
accommodations are necessary, you may register with Accessibility
Services at (404)727-9877 or via e-mail at
\href{mailto:accessibility@emory.edu}{\nolinkurl{accessibility@emory.edu}}.
To register with OAS, students must self identify and initiate contact
with the OAS office.

\hypertarget{absence-policy}{%
\subsection{Absence Policy}\label{absence-policy}}

Missing 25\% or more of class meetings will result in automatic failure
of a course. In other words, students absent seven (7) or more times, in
a course that meets twice a week, will receive a grade of ``F'' for the
course. Absences include trips, appointments, interviews, conferences,
illness, injury, as well as simply not showing up. Religious
observances, school business, and major illness will be considered;
however, notify me in advance of any planned absences and submit your
assignment prior to the event. After any absence, it is your
responsibility to find out what material, assignments, or announcements
you missed.

\hypertarget{laurens-promise}{%
\subsection{Lauren's Promise}\label{laurens-promise}}

Lauren McCluskey, a 21-year-old honors student athlete,
\href{https://www.sltrib.com/opinion/commentary/2019/02/10/commentary-failing-lauren/}{was
murdered on October 22, 2018 by a man she briefly dated on the
University of Utah campus}. We must all take action to ensure that this
never happens again. If you are in immediate danger, call 911 or Emory
police (404-727-6111). \textbf{I will listen and believe you if someone
is threatening you.}

Any form of sexual harassment or violence will not be excused or
tolerated at Emory. If you are experiencing sexual assault, domestic
violence, or stalking, please report it to me or directly to
\href{http://respect.emory.edu/index.html}{Emory's Office of Respect}
(470-270-5360).

\hypertarget{assignments-and-grading}{%
\section{Assignments and Grading}\label{assignments-and-grading}}

Your final grade consists of five homework assignments and a final class
project/paper, along with a participation component. I describe each
assignment below, with more additional detail provided on our
\href{https://econ470s23.classes.ianmccarthyecon.com/}{class website}.

\hypertarget{homework-assignments}{%
\subsection{Homework assignments}\label{homework-assignments}}

There are 5 homework assignments in this class (one for each module).
Details of the assignments are on our
\href{https://econ470s23.classes.ianmccarthyecon.com/}{class website}.
Each assignment will have 10 questions and will be worth 30 points
total. Each question is worth 3 points, allocated as follows:

\begin{itemize}
\tightlist
\item
  1 point for an attempted answer as of the draft due date
\item
  1 point for an attempted answer as of the revision due date
\item
  1 point for a correct answer as of the final due date
\end{itemize}

Notice that each assignment has three different stages: a draft stage, a
revision stage, and a final stage. The draft stage reflects your
original attempt. In the revision stage, we will discuss the homework in
class, break into small groups to share feedback and questions, and
revise the homework. In the final stage, we will repeat this process one
last time with any final questions. Updated assignments are due at each
stage.

The homework assignments are almost entirely empirical, with each
assignment focusing on a specific identification strategy, research
question, and health-related dataset. I will provide all answers in
\textbf{R}; however, you are free to use any software you'd like. Just
keep in mind that I will be best able to answer your questions if you
also use \textbf{R}. I can also answer most any question if you choose
to use \textbf{Stata}. I can point you in the right direction with
\textbf{Python}, but I likely can't provide you with any sample code or
specific syntax advice.

Due dates for each stage are listed below:

\begin{itemize}
\tightlist
\item
  Homework 1: Initial date (1/23), revision date (1/25), final date
  (1/30)
\item
  Homework 2: Initial date (2/13), revision date (2/15), final date
  (2/20)
\item
  Homework 3: Initial date (3/13), revision date (3/15), final date
  (3/20)
\item
  Homework 4: Initial date (4/3), revision date (4/5), final date (4/10)
\item
  Homework 5: Initial date (4/17), revision date (4/19), final date
  (4/24)
\end{itemize}

\hypertarget{participation}{%
\subsection{Participation}\label{participation}}

Participation is graded based on your engagement in small-group homework
discussions. A large portion of this class is about learning how to work
with \textbf{real world} data. To do this, we will spend significant
in-class time reviewing the assignments and updating our work based on
collective feedback. For each assignment, students will be placed into
small groups to discuss each others' work. Your participation grade will
derive from a peer evaluation based on these small group sessions.

At the end of each assignment, each member of each group will be asked
to rate the feedback of their peers on a scale of 1-3. A value of 1
reflects insufficient effort or engagement, a value of 2 reflects
sufficient effort, and a value of 3 reflects excellent effort. I
anticipate small groups of at least 4 students, which means you will
receive peer scores from at least 3 other students for each assignment.
Your median score across all peer evaluations will be used as your
participation grade in that assignment. For example, if you receive peer
scores of 3, 3, and 1, your participation score would be 3. If the
median value is a tie, your participation grade will be the highest of
the possible values.

At the end of the semester, your peer grades from each assignment will
be totaled. Your final participation grade will be based on your total
number of participation points out of 10. Note there are 15 possible
participation points, so there is some margin for error here if you
happen to miss class during a small group session or simply aren't
engaged in one of the review sessions for some reason.

\hypertarget{final-project}{%
\subsection{Final project}\label{final-project}}

We do not have a final exam in this class. Instead, we have a final
project that asks you to submit a 5-page policy brief. Please see the
\href{https://econ470s23.classes.ianmccarthyecon.com/}{class website}
for more details. The policy brief is worth 90 points (36\%) toward your
final grade.

\hypertarget{due-dates}{%
\subsection{Due dates}\label{due-dates}}

This section is just to highlight the final due dates for each
assignment. Late assignments will receive an automatic 2\% reduction for
each day the assignment is turned in after the due date.

\begin{itemize}
\tightlist
\item
  \textbf{January 30}: Homework 1
\item
  \textbf{February 20}: Homework 2
\item
  \textbf{March 20}: Homework 3
\item
  \textbf{April 10}: Homework 4
\item
  \textbf{April 24}: Homework 5
\item
  \textbf{TBD}: Final Project
\end{itemize}

\hypertarget{grades}{%
\subsection{Grades}\label{grades}}

Each assignment will contribute to your final grade as described below.

\begin{longtable}[]{@{}lrl@{}}
\toprule()
Assignment & Points & Percent \\
\midrule()
\endhead
Participation & 10 & 4\% \\
Homework (30 x 5) & 150 & 60\% \\
Final project & 90 & 36\% \\
\bottomrule()
\end{longtable}

Your final percentage grade comes from your total points as a percent of
all possible points available in the class (250). That percent then
translates to a letter grade as follows:

\begin{longtable}[]{@{}llll@{}}
\toprule()
Grade & Range & Grade & Range \\
\midrule()
\endhead
A & 93-100\% & C & 73-76\% \\
A- & 90-92\% & C- & 70-72\% \\
B+ & 87-89\% & D+ & 67-69\% \\
B & 83-86\% & D & 60-66\% \\
B- & 80-82\% & F & \textless{} 60\% \\
C+ & 77-79\% & & \\
\bottomrule()
\end{longtable}

\hypertarget{class-schedule}{%
\section{Class Schedule}\label{class-schedule}}

Below is a preliminary outline (subject to change) of specific topics
and resources throughout the semester. Please check the class website
for the most up-to-date schedule as well as links to specific resources
for each week.

\hypertarget{module-0-introduction}{%
\subsection{Module 0: Introduction}\label{module-0-introduction}}

\hypertarget{motivation-and-introduction-to-version-control}{%
\subsubsection{1/11: Motivation and introduction to version
control}\label{motivation-and-introduction-to-version-control}}

\begin{itemize}
\tightlist
\item
  Lesson 2 from \href{https://github.com/uo-ec607/lectures}{Data Science
  for Economists}. This is part of a course from
  \href{https://grantmcdermott.com/}{Grant McDermott} with a ton of
  great online resources.
\item
  \href{https://r4ds.had.co.nz/}{R for Data Science}
\end{itemize}

\hypertarget{introduction-to-causal-inference}{%
\subsubsection{1/16, 1/18: Introduction to causal
inference}\label{introduction-to-causal-inference}}

\begin{itemize}
\tightlist
\item
  Chapters 1-2, and 4 of \href{https://mixtape.scunning.com/}{Causal
  Inference: The Mixtape}
\item
  Chapters 1-5, 10, and 13 of \href{https://theeffectbook.net/}{The
  Effect: An Introduction to Research Design and Causality}
\item
  Chapters 2-3 of Angrist and Pischke, \emph{Mostly {Harmless}
  {Econometrics}}.
\end{itemize}

\hypertarget{submission-and-revision-of-homework-1-version-control-and-data-management}{%
\subsubsection{1/23, 1/25: Submission and revision of Homework 1
(Version Control and Data
Management)}\label{submission-and-revision-of-homework-1-version-control-and-data-management}}

\hypertarget{module-1-pricing-and-selection-on-observables}{%
\subsection{Module 1: Pricing and Selection on
Observables}\label{module-1-pricing-and-selection-on-observables}}

\hypertarget{hospital-prices-and-hcris-data}{%
\subsubsection{1/30, 2/1: Hospital prices and HCRIS
data}\label{hospital-prices-and-hcris-data}}

\begin{itemize}
\tightlist
\item
  \href{https://github.com/imccart/HCRIS}{Hospital Cost Report
  Information System (HCRIS)} repository
\item
  See also a similar repository
  \href{https://github.com/asacarny/hospital-cost-reports}{here} from
  \href{http://sacarny.com/}{Adam Sacarny}
\end{itemize}

\hypertarget{selection-on-observables}{%
\subsubsection{2/6, 2/8: Selection on
observables}\label{selection-on-observables}}

\begin{itemize}
\tightlist
\item
  Chapter 5 of \href{https://mixtape.scunning.com/}{Causal Inference:
  The Mixtape}
\item
  Chapter 14 of \href{https://theeffectbook.net/}{The Effect: An
  Introduction to Research Design and Causality}
\end{itemize}

\hypertarget{submission-and-revision-of-homework-2-hospital-prices-and-selection-on-observables}{%
\subsubsection{2/13, 2/15: Submission and revision of Homework 2
(Hospital Prices and Selection on
Observables)}\label{submission-and-revision-of-homework-2-hospital-prices-and-selection-on-observables}}

\hypertarget{module-2-demand-for-cigarettes-and-instrumental-variables}{%
\subsection{Module 2: Demand for Cigarettes and Instrumental
Variables}\label{module-2-demand-for-cigarettes-and-instrumental-variables}}

\hypertarget{smoking-and-cdc-data}{%
\subsubsection{2/20, 2/22: Smoking and CDC
data}\label{smoking-and-cdc-data}}

\begin{itemize}
\tightlist
\item
  \href{https://github.com/imccart/CDC-Tobacco}{CDC Tax Burden on
  Tobacco}\\
\item
  Initial submission and revision of Homework 2
\end{itemize}

\hypertarget{instrumental-variables}{%
\subsubsection{2/27, 3/1: Instrumental
variables}\label{instrumental-variables}}

\begin{itemize}
\tightlist
\item
  Chapter 7 of \href{https://mixtape.scunning.com/}{Causal Inference:
  The Mixtape}
\item
  Chapter 19 of \href{https://theeffectbook.net/}{The Effect: An
  Introduction to Research Design and Causality}
\end{itemize}

\hypertarget{submission-and-revision-of-homework-3-demand-for-cigarettes-and-iv}{%
\subsubsection{3/13, 3/15: Submission and revision of Homework 3 (Demand
for Cigarettes and
IV)}\label{submission-and-revision-of-homework-3-demand-for-cigarettes-and-iv}}

\hypertarget{module-3-medicare-advantage-and-regression-discontinuity}{%
\subsection{Module 3: Medicare Advantage and Regression
Discontinuity}\label{module-3-medicare-advantage-and-regression-discontinuity}}

\hypertarget{medicare-advantage}{%
\subsubsection{3/20, 3/22: Medicare
Advantage}\label{medicare-advantage}}

\begin{itemize}
\tightlist
\item
  \href{https://github.com/imccart/Medicare-Advantage}{Medicare
  Advantage} repository
\end{itemize}

\hypertarget{regression-discontinuity}{%
\subsubsection{3/27, 3/29: Regression
discontinuity}\label{regression-discontinuity}}

\begin{itemize}
\tightlist
\item
  Chapter 6 of \href{https://mixtape.scunning.com/}{Causal Inference:
  The Mixtape}
\item
  Chapter 20 of \href{https://theeffectbook.net/}{The Effect: An
  Introduction to Research Design and Causality}
\end{itemize}

\hypertarget{submission-and-revision-of-homework-4-quality-ratings-insurance-choice-and-rd}{%
\subsubsection{4/3, 4/5: Submission and revision of Homework 4 (Quality
Ratings, Insurance Choice, and
RD)}\label{submission-and-revision-of-homework-4-quality-ratings-insurance-choice-and-rd}}

\hypertarget{module-4-medicaid-expansion-and-difference-in-differences}{%
\subsection{Module 4: Medicaid Expansion and
Difference-in-Differences}\label{module-4-medicaid-expansion-and-difference-in-differences}}

\hypertarget{difference-in-differences}{%
\subsubsection{4/10, 4/12:
Difference-in-Differences}\label{difference-in-differences}}

\begin{itemize}
\tightlist
\item
  \href{https://github.com/imccart/Insurance-Access}{Insurance Access
  and Medicaid Expansion} repository
\item
  Chapters 8-9 of \href{https://mixtape.scunning.com/}{Causal Inference:
  The Mixtape}
\item
  Chapters 16-18 of \href{https://theeffectbook.net/}{The Effect: An
  Introduction to Research Design and Causality}
\end{itemize}

\hypertarget{submission-and-revision-of-homework-5-medicaid-expansion-uninsurance-and-did}{%
\subsubsection{4/17, 4/19: Submission and revision of Homework 5
(Medicaid Expansion, Uninsurance, and
DiD)}\label{submission-and-revision-of-homework-5-medicaid-expansion-uninsurance-and-did}}

\hypertarget{discussion-of-final-policy-briefs}{%
\subsubsection{4/24: Discussion of final policy
briefs}\label{discussion-of-final-policy-briefs}}




\end{document}

\makeatletter
\def\@maketitle{%
  \newpage
%  \null
%  \vskip 2em%
%  \begin{center}%
  \let \footnote \thanks
    {\fontsize{18}{20}\selectfont\raggedright  \setlength{\parindent}{0pt} \@title \par}%
}
%\fi
\makeatother