---
title: "Module 2: Demand for Cigarettes and Instrumental Variables"
subtitle: "Part 3: Application of IV to Demand Estimation"
author: Ian McCarthy | Emory University
date: Econ 470 & HLTH 470 #"`r format(Sys.time(), '%d %B %Y')`"
header-includes: 
  - \usepackage{graphicx}
  - \usepackage{amsmath}
output:
#  html_document: default  
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, custom.css] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "macros.js"
knit: pagedown::chrome_print   
---

<!-- Adjust some CSS code for font size and maintain R code font size -->
<style type="text/css">
.remark-slide-content {
    font-size: 30px;
    padding: 1em 2em 1em 2em;    
}
.remark-code {
  font-size: 15px;
}
.remark-inline-code { 
    font-size: 20px;
}
</style>


<!-- Set R options for how code chunks are displayed and load packages -->
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
knitr::opts_chunk$set(
  fig.align="center",  
  fig.height=3, #fig.width=6,
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=T,# echo=F, warning=F, message=F
  warning = FALSE, 
  message = FALSE, 
  cache.lazy = FALSE,
  error=TRUE
  )

knitr::opts_hooks$set(fig.callout = function(options) {
  if(options$fig.callout) {
    options$echo = FALSE
  }
  options
})

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, readr, readxl, hrbrthemes,
               scales, plotly, gganimate, cobalt, ivpack, stargazer, ggthemes, here)
```

```{r xaringan-scribble, echo=FALSE}
xaringanExtra::use_scribble(rgb(0.9, 0.5, 0.5))
```


# Naive estimate
Clearly a strong relationship between prices and sales. For example, just from OLS:
```{r echo=FALSE}
cig.data <- read_rds(here("data/TaxBurden_Data.rds"))
cig.data <- cig.data %>% mutate(ln_sales=log(sales_per_capita),
                                ln_price_cpi=log(price_cpi),
                                ln_price=log(cost_per_pack),
                                tax_cpi=tax_state*(218/index),
                                total_tax_cpi=tax_dollar*(218/index),                                
                                ln_state_tax=log(tax_cpi))
ols <- lm(ln_sales ~ ln_price, data=cig.data)
summary(ols)
```

---
# Is this causal?
- But is that the true demand curve?

- Aren't other things changing that tend to reduce cigarette sales?

---
# Tax as an IV
```{r cig-tax, eval=FALSE, warning=FALSE}
cig.data %>% 
  ggplot(aes(x=Year,y=total_tax_cpi)) + 
  stat_summary(fun.y="mean",geom="line") +
  labs(
    x="Year",
    y="Tax per Pack ($)",
    title="Cigarette Taxes in 2010 Real Dollars"
  ) + theme_bw() +
  scale_x_continuous(breaks=seq(1970, 2020, 5))
```

.plot-callout[
```{r cig-tax-callout, ref.label="cig-tax", fig.callout=TRUE, warning=FALSE}
```
]

---
# Tax as an IV

```{r cig-tax-output, ref.label="cig-tax", fig.callout=TRUE, warning=FALSE}
```


---
# IV Results

```{r echo=FALSE}
ivs <- ivreg(ln_sales ~ ln_price | total_tax_cpi, 
             data=cig.data)
summary(ivs)
```

---
# Two-stage equivalence
```{r}
step1 <- lm(ln_price ~ total_tax_cpi, data=cig.data)
pricehat <- predict(step1)
step2 <- lm(ln_sales ~ pricehat, data=cig.data)
summary(step2)
```




---
# Different specifications
```{r echo=FALSE, results="asis"}
ols1 <- lm(ln_sales ~ ln_price_cpi, data=cig.data)
ols2 <- lm(ln_sales ~ ln_price_cpi + factor(state), data=cig.data)
ols3 <- lm(ln_sales ~ ln_price_cpi + factor(state) + factor(Year), data=cig.data)

ivs1 <- ivreg(ln_sales ~ ln_price_cpi | total_tax_cpi, data=cig.data)
ivs2 <- ivreg(ln_sales ~ ln_price_cpi + factor(state) | 
                total_tax_cpi + factor(state), data=cig.data)
ivs3 <- ivreg(ln_sales ~ ln_price_cpi + factor(state) + factor(Year) | 
                total_tax_cpi + factor(state) + factor(Year), data=cig.data)
stargazer(ols1, ols2, ols3, ivs1, ivs2, ivs3, 
          keep=c("ln_price_cpi"), 
          type="html",
          column.labels = c("OLS","IV"),
          column.separate=c(3,3),
          keep.stat=c("n"),
          model.names=FALSE,
          dep.var.caption="",
          font.size="footnotesize",
          no.space=TRUE,
          add.lines=list(c("State FE","No","Yes","Yes","No","Yes","Yes"),
                         c("Year FE","No","No","Yes","No","No","Yes")),
          covariate.labels = "Log Price",
          dep.var.labels = "Log Sales per Capita")
```


---
# Test the IV
```{r echo=FALSE, results="asis"}
first1 <- lm(ln_price_cpi ~ total_tax_cpi, data=cig.data)
first2 <- lm(ln_price_cpi ~ total_tax_cpi + factor(state), data=cig.data)
first3 <- lm(ln_price_cpi ~ total_tax_cpi + factor(state) + factor(Year), data=cig.data)

rf1 <- lm(ln_sales ~ total_tax_cpi, data=cig.data)
rf2 <- lm(ln_sales ~ total_tax_cpi + factor(state), data=cig.data)
rf3 <- lm(ln_sales ~ total_tax_cpi + factor(state) + factor(Year), data=cig.data)


stargazer(first1, first2, first3, rf1, rf2, rf3,
          keep=c("total_tax_cpi"), 
          type="html",
          column.labels = c("First Stage","Reduced Form"),
          column.separate=c(3,3),
          keep.stat=c("n"),
          model.names=FALSE,
          dep.var.caption="",
          font.size="footnotesize",
          no.space=TRUE,
          add.lines=list(c("State FE","No","Yes","Yes","No","Yes","Yes"),
                         c("Year FE","No","No","Yes","No","No","Yes")),
          covariate.labels = "Tax per Pack",
          dep.var.labels = c("Log Price","Log Sales"))
```

---
# Summary
1. Most elasticities of around -0.25% to -0.37%

2. Much larger elasticities when including year fixed effects

3. Perhaps not too outlandish given more recent evidence: [NBER Working Paper](https://www.nber.org/papers/w22251.pdf).

---
# Some other IV issues

1. IV estimators are biased. Performance in finite samples is questionable.

2. IV estimators provide an estimate of a Local Average Treatment Effect (LATE), which is only the same as the ATT under some conditions or assumptions.

3. What about lots of instruments? The finite sample problem is more important and we may try other things (JIVE).<br>

--
<br>
The National Bureau of Economic Researh (NBER) has a great resource [here](https://www.nber.org/econometrics_minicourse_2018/2018si_methods.pdf) for understanding instruments in practice.


---
# Quick IV Review

1. When do we consider IV as a potential identification strategy?

2. What are the main IV assumptions (and what do they mean)?

3. How do we test for those assumptions?
