---
title: "Module 3: Medicare Advantage Quality and Regression Discontinuity"
subtitle: "Part 1: MA Data and Quality Measurement"
author: Ian McCarthy | Emory University
date: Econ 470 & HLTH 470 #"`r format(Sys.time(), '%d %B %Y')`"
header-includes: 
  - \usepackage{graphicx}
  - \usepackage{amsmath}
output:
#  html_document: default
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, custom.css] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "macros.js"
knit: pagedown::chrome_print
---

<!-- Adjust some CSS code for font size and maintain R code font size -->
<style type="text/css">
.remark-slide-content {
    font-size: 30px;
    padding: 1em 2em 1em 2em;    
}
.remark-code {
  font-size: 15px;
}
.remark-inline-code { 
    font-size: 20px;
}
</style>


<!-- Set R options for how code chunks are displayed and load packages -->
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
knitr::opts_chunk$set(
  fig.align="center",  
  fig.height=3, #fig.width=6,
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=T,# echo=F, warning=F, message=F
  warning = FALSE, 
  message = FALSE, 
  cache.lazy=FALSE,
  error=TRUE
  )

knitr::opts_hooks$set(fig.callout = function(options) {
  if(options$fig.callout) {
    options$echo = FALSE
  }
  options
})

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, readr, readxl, hrbrthemes,
               scales, plotly, gganimate, cobalt, ivpack, stargazer, haven, ggthemes,
               gifski, magick, rdrobust, here)
set.seed(1234)
```

```{r xaringan-scribble, echo=FALSE}
xaringanExtra::use_scribble(rgb(0.9, 0.5, 0.5))
```


# What is Medicare Advantage
- Private provision of health insurance for Medicare beneficiaries
- Medicare "replacement" plans <br>

--
- It's just private insurance for Medicare folks

---
# Medicare Advantage History
- Existed since 1980s, formalized in the 1990s, expanded in 2000s
- Medicare+Choice as part of Balanced Budget Act in 1997
- Largest expansion: Medicare Modernization Act in 2003 (also brought Medicare Part D)

---
# Medicare Advantage Details
In its current form...
- Insurers submit plan details and a price needed to cover traditional Medicare ("bid")
- If approved, Medicare pays risk-adjusted bid *or* benchmark
- Bid $<$ benchmark, insurer gets a rebate
- Bid $>$ benchmark, insurer charges premium
- Seperate bidding for Part D

---
# Medicare Advantage in Real Life
Let's take a look at the Medicare Advantage plan options...

--
<br>
[Medicare Plan Finder](https://www.medicare.gov/plan-compare/)


---
# Medicare Advantage Quality Ratings
- Initial MA Star Ratings (2007)
- Overall rating introduced in 2009
- Complicated formula...<br>

--
<br>

<b>key point:</b> ratings from several domains are averaged and then rounded



# Medicare Advantage Data
- Recall the Medicare Advantage repository, [Medicare Advantage GitHub repository](https://github.com/imccart-test/ma)

--
<br>

- Now we need to work with the **final** dataset

---
# Final MA Data
```{r eval=T, include=T}
ma.data <- read_rds(here("data/final_ma_data.rds"))
```


---
# Summary stats
Focus on enrollments and star ratings:
```{r, results="asis"}
stargazer(as.data.frame(ma.data %>% select(avg_enrollment, avg_eligibles, Star_Rating)), type="html")
```


---
# Clean the data
Limit to plans with:
- Observed enrollments, $>$ 10
- First year of star rating (2009)
- Observed star rating

--
<br>

```{r}
ma.data.clean <- ma.data %>%
  filter(!is.na(avg_enrollment) & year==2009 & !is.na(partc_score)) #<<
```

---
# Calculate raw average rating
```{r}
ma.data.clean <- ma.data.clean %>%
  mutate(raw_rating=rowMeans(
    cbind(breastcancer_screen,rectalcancer_screen,cv_cholscreen,diabetes_cholscreen,
          glaucoma_test,monitoring,flu_vaccine,pn_vaccine,physical_health,
          mental_health,osteo_test,physical_monitor,primaryaccess,
          hospital_followup,depression_followup,nodelays,carequickly,
          overallrating_care,overallrating_plan,calltime,
          doctor_communicate,customer_service,osteo_manage,
          diabetes_eye,diabetes_kidney,diabetes_bloodsugar,
          diabetes_chol,antidepressant,bloodpressure,ra_manage,
          copd_test,betablocker,bladder,falling,appeals_timely,
          appeals_review),
    na.rm=T)) %>%
  select(contractid, planid, fips, avg_enrollment, first_enrollment,
         last_enrollment, state, county, raw_rating, partc_score,
         avg_eligibles, avg_enrolled, premium_partc, risk_ab, Star_Rating,
         bid, avg_ffscost, ma_rate)
```

